---
title: "VAST index workflow"
author: "Sarah Gaichas and Sean Lucey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(here)
library(DT)
```

# Introduction

We used stomach contents data to estimate an index of forage fish biomass for the 2023 state of the ecosystem reports and the bluefish stock assessment. Here, we extend that idea to estimate an index of benthic group biomass for use in fitting food web models of the Gulf of Maine (GOM) and George Bank (GB). 

The basic workflow is to develop a dataset of stomach contents data where fish predators act as samplers of the prey field, then fit a vector autoregressive spatio-temporal (VAST) model to this dataset to generate an index. 

# Data

## Which prey?

Sean has mapped the food habits categories to Rpath categories. We want indices for the benthic categories Megafauna and Macrofauna in the GB and GOM Rpath models.

Upon review of these categories and model input data by Sean and Sarah Weisberg, two changes were made from the original list.

*  The prey "PECTINIDAE", "PECTINIDAE SHELL", "PECTINIDAE VISCERA" were removed from macrobenthos and added to the AtlScallop group, as they likely represent sea scallops identified in stomachs. Therefore these prey categories will not be included in the benthos indices.

*  The prey "ALBUNEA PARETII", "EMERITA TALPOIDA", "RANILIA MURICATA", "HIPPIDAE" were removed from megabenthos and added to macrobenthos, because these are small sand crabs and mole crabs.

All tables in this document have been adjusted to reflect these changes.


```{r}
# object is called prey
load(here("data/prey.RData")) #original mappings from June 2023

addtomacro <- prey |>
  dplyr::filter(RPATH == "Megabenthos") |> 
  dplyr::filter(PYNAM %in% c("ALBUNEA PARETII", "EMERITA TALPOIDA", "RANILIA MURICATA", "HIPPIDAE"))

megaben <- prey |>
  dplyr::filter(RPATH == "Megabenthos") |>
  dplyr::filter(!PYNAM %in% c("ALBUNEA PARETII", "EMERITA TALPOIDA", "RANILIA MURICATA", "HIPPIDAE"))

macroben <- prey |>
  dplyr::filter(RPATH == "Macrobenthos") |> 
  dplyr::filter(!PYNAM %in% c("PECTINIDAE", "PECTINIDAE SHELL", "PECTINIDAE VISCERA")) |>
  dplyr::bind_rows(addtomacro)


```


The Macrobenthos Rpath category has `r dim(macroben)[1]` food habits database species codes:

```{r}
datatable(macroben, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Macrobenthos species names and codes",
          options = list(pageLength = 10,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))
```

The Megabenthos Rpath category has `r dim(megaben)[1]` food habits database species codes:

```{r}
datatable(megaben, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Megabenthos species names and codes",
          options = list(pageLength = 10,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))
```

Which of these show up most often in predator stomachs?

```{r}
# object is called `allfh`
load(url("https://github.com/NOAA-EDAB/forageindex/raw/main/fhdat/allfh.Rdata"))

#object is called allfh21
load(url("https://github.com/NOAA-EDAB/forageindex/raw/main/fhdat/allfh21.RData"))

allfh <- allfh %>%
  dplyr::bind_rows(allfh21)

macrobenfh <- allfh %>%
  dplyr::left_join(macroben %>% 
                     dplyr::select(PYNAM, RPATH) %>%
                     setNames(., tolower(names(.)))) %>%
  dplyr::filter(!is.na(rpath))

megabenfh <- allfh %>%
  dplyr::left_join(megaben %>% 
                     dplyr::select(PYNAM, RPATH) %>%
                     setNames(., tolower(names(.)))) %>%
  dplyr::filter(!is.na(rpath))

macropreycount <- macrobenfh %>%
   #group_by(pdcomnam, pynam) %>%
   group_by(pynam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))
   #pivot_wider(names_from = pdcomnam, values_from = count) 

megapreycount <- megabenfh %>%
   #group_by(pdcomnam, pynam) %>%
   group_by(pynam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))
   #pivot_wider(names_from = pdcomnam, values_from = count) 

datatable(macropreycount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of stomachs with each macrobenthos species across all predators, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

datatable(megapreycount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of stomachs with each megabenthos species across all predators, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

```


## Which predators?

We have a couple of options for a predator list. First is to use every predator (or all above a cutoff number of observations, maybe 50 or 100 across the whole database) where something in the prey list was observed, second is to use all benthivores or some other group(s) as identified by diet similarity.

```{r}

macropredcount <- macrobenfh %>%
  group_by(pdcomnam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))

megapredcount <- megabenfh %>%
  group_by(pdcomnam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))

datatable(macropredcount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of macrobenthos species observations by predator, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

datatable(megapredcount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of megabenthos species observations by predator, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))



```

The clusters based on diet similarity show several clusters that appear on the lists above. I think because we want to include so many prey groups, the cluster groupings might actually limit us, or we would have to use more than one cluster:

```{r, fig.height=12}
dietoverlap <- read_csv(here("data-raw/tgmat.2022-02-15.csv"))

# follows example here https://cran.r-project.org/web/packages/dendextend/vignettes/Cluster_Analysis.html

library(dendextend)

d_dietoverlap <- dist(dietoverlap)

guilds <- hclust(d_dietoverlap)

#plot(guilds)

dend <- as.dendrogram(guilds)

dend <- rotate(dend, 1:136)

dend <- color_branches(dend, k=6) # Brian uses 6 categories

labels(dend) <- paste(as.character(names(dietoverlap[-1]))[order.dendrogram(dend)],
                           "(",labels(dend),")", 
                           sep = "")

dend <- hang.dendrogram(dend,hang_height=0.1)

# reduce the size of the labels:
# dend <- assign_values_to_leaves_nodePar(dend, 0.5, "lab.cex")
dend <- set(dend, "labels_cex", 0.5)
# And plot:
par(mar = c(3,3,3,7))
plot(dend, 
     main = "Clustered NEFSC diet data, (complete)
     (the labels give the predator species/size)", 
     horiz =  TRUE,  nodePar = list(cex = .007))
#legend("topleft", legend = iris_species, fill = rainbow_hcl(3))

```

We decided to use the cluster groups to eliminate pelagic species that would only rarely consume these prey. Therefore we will use species in 4 of the 6 clusters, but not use species in the planktivore cluster (blue above) or the piscivore cluster (gold above). 

Therefore, the following is the NEFSC BTS predator/size list:

```{r}
pisc <- partition_leaves(dend)[[
  which_node(dend, c("Bluefish..S(37)", "Bluefish..M(36)", "Bluefish..L(35)"))
]]

piscdf <- data.frame("COMNAME" = toupper(str_remove(pisc, "\\..*")),
                             "SizeCat" = str_remove(str_extract(pisc, "\\..*[:upper:]+"), "\\.."))


plank <- partition_leaves(dend)[[
  which_node(dend, c("Atlantic herring..S(20)", "Atlantic mackerel..S(23)", "Blueback herring..XS(34)"))
]]

plankdf <- data.frame("COMNAME" = toupper(str_remove(plank, "\\..*")),
                             "SizeCat" = str_remove(str_extract(plank, "\\..*[:upper:]+"), "\\.."))

all <- partition_leaves(dend)[[
  which_node(dend, c("Barndoor skate..S(26)", "Bluefish..L(35)"))
]]

alldf <- data.frame("COMNAME" = toupper(str_remove(all, "\\..*")),
                             "SizeCat" = str_remove(str_extract(all, "\\..*[:upper:]+"), "\\.."))

sizedef <- readxl::read_excel(here::here("data-raw/Table3.xlsx"),
                              range = "B3:H55") |>  # table of NEFSC FH size categories from Brian
  dplyr::rename(XS = `Extra-Small`,
                S = "Small",
                M = "Medium",
                L = "Large",
                XL = `Extra-Large`,
                CommonName = `Common Name`,
                SpeciesName = `Species Name`) |>
  tidyr::pivot_longer(-c(1:2), names_to = "sizecat", values_to = "range") |>
  dplyr::filter(!range=="-") |>
  dplyr::mutate(COMNAME = toupper(CommonName)) 
  

# add missing by hand from word Table 1.docx 
# (pollock was misspelled, fixed in spreadsheet)
# (removed "flounder" from Windowpane in spreadsheet)
# asked Brian for these; response:
# For these predators and most likely any others not listed in TM216, 
# the default is S (<=20 cm), M (>20 and <=50 cm), and L (>50 cm).
# Northern kingfish M
# Buckler dory S
# Fourbeard rockling S
# Fourbeard rockling M
# Cunner M

extra <- data.frame(CommonName = c("Northern kingfish",
                                      "Buckler dory",
                                      "Fourbeard rockling",
                                      "Fourbeard rockling",
                                      "Cunner"),  
                    sizecat = c("M",
                                "S",
                                "S",
                                "M",
                                "M"), 
                    range = c("21-50",
                              "≤20",
                              "≤20",
                              "21-50",
                              "21-50"))  |>
  dplyr::mutate(COMNAME = toupper(CommonName)) 

sizedef <- bind_rows(sizedef, extra)

 # dplyr::mutate(COMNAME = toupper(`Common Name`),
 #               XS = readr::parse_number(`Extra-Small`),
 #               S = readr::parse_number(Small),
 #               M = readr::parse_number(Medium),
 #               L = readr::parse_number(Large),
 #               XL = readr::parse_number(`Extra-Large`))
  
  
# we'll call benthivores everything but piscivores and planktivores
benthivores <- alldf |>
  dplyr::anti_join(dplyr::bind_rows(piscdf, plankdf)) |>
  dplyr::left_join(sizedef, join_by(COMNAME == COMNAME, SizeCat == sizecat)) |>
  dplyr::left_join(ecodata::species_groupings, by = "COMNAME") |>
  dplyr::select(COMNAME, ITISSPP, SCINAME, SizeCat = SizeCat.x, SizeRange_cm = range) |>
  dplyr::distinct()

datatable(benthivores, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Predator list for benthos indices",
          options = list(pageLength = 90,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

saveRDS(benthivores, here::here("data/benthivorelist.rds"))

```


## Which surveys?

Start with NEFSC, find out if MA and ME/NH surveys collect diet information, and add NEAMAP later once predator and prey lists are well established.

MA and ME/NH surveys do not collect diet information.

As of January 2024, predator and prey lists are established enough for a NEAMAP request. Just needed to add the size categories; almost there...

Add ITISSPP to prey lists where possible

```{r}
macrobenITIS <- macroben |>
  dplyr::select(PYNAM, PYCOMNAM, PYSPP) |>
  dplyr::left_join(ecodata::species_groupings, join_by(PYNAM == SCINAME)) |>
  dplyr::select(PYNAM, PYCOMNAM, COMNAME, PYSPP, ITISSPP)

megabenITIS <- megaben |>
  dplyr::select(PYNAM, PYCOMNAM, PYSPP) |>
  dplyr::left_join(ecodata::species_groupings, join_by(PYNAM == SCINAME)) |>
  dplyr::select(PYNAM, PYCOMNAM, COMNAME, PYSPP, ITISSPP)

saveRDS(macrobenITIS, here::here("data/macrobenthosprey.rds"))
saveRDS(megabenITIS, here::here("data/megabenthosprey.rds"))

```

NEAMAP survey results received May 2024 using the above prey lists. 

Jim Gartland provided a lookup table of NEFSC to NEAMAP prey codes! Add to ecodata for future use.



All data combined in the script VASTbenthos_ProcessInputDat.R 

```{r}

```

For this modify the data generation script in the forageindex repo, here: https://github.com/NOAA-EDAB/forageindex/blob/main/fhdat/VASTforage_ProcessInputDat.R 

## Spatial scale

Suggestion: use entire VAST northwest-atlantic grid for estimation and split out GOM and GB as was done for the forage fish index.

## Covariates

Suggestion: start with mean predator size at a station, number of predator species at a station, and bottom water temperature. Is there anything else we would want for benthic organisms? Reviewers suggested depth for the forage fish index, but it did not converge and may be more appropriate as a habitat covariate. 

### Fill missing physical data?

When I looked at bottom temperature data in the NEFSC survey, we had as many missing values as for surface temperature, so it is likely we will want to fill these with other data sources to avoid losing information. Hubert DuPontavice provided his reconstruction of bottom temperatures based on GLORYS and ROMS outputs for the NEUS. The full 1.1 GB dataset is stored locally in data-raw/bottomtemp and the identical source file is [here on google drive](https://drive.google.com/file/d/11HanY5Tzu--77HevCV5gRdNvaAvuYVYU/view?usp=share_link).

# Modeling

## VAST model setup

Univariate or multivariate? First lets try separate univariate models for each of the prey groups, one model for macrobenthos and one for megabenthos. Then we can get fancier trying to look at a multivariate model?

Do we want to separate seasons or try an annual model? I guess it depends on how the Rpaths are set up.

DECISION: Fall is how the GOM is parameterized so start with that.

## VAST model selection

## VAST model bias correction and results

