---
title: "VAST index workflow"
author: "Sarah Gaichas and Sean Lucey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(here)
library(DT)
```

# Introduction

We used stomach contents data to estimate an index of forage fish biomass for the 2023 state of the ecosystem reports and the bluefish stock assessment. Here, we extend that idea to estimate an index of benthic group biomass for use in fitting food web models of the Gulf of Maine (GOM) and George Bank (GB). 

The basic workflow is to develop a dataset of stomach contents data where fish predators act as samplers of the prey field, then fit a vector autoregressive spatio-temporal (VAST) model to this dataset to generate an index. 

# Data

## Which prey?

Sean has mapped the food habits categories to Rpath categories. We want indices for the benthic categories Megafauna and Macrofauna in the GB and GOM Rpath models:

```{r}
# object is called prey
load(here("data/prey.Rdata"))

macroben <- prey %>%
  dplyr::filter(RPATH == "Macrobenthos")

megaben <- prey %>%
  dplyr::filter(RPATH == "Megabenthos")

```

The Macrobenthos Rpath category has `r dim(macroben)[1]` food habits database species codes:

```{r}
datatable(macroben, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Macrobenthos species names and codes",
          options = list(pageLength = 10,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))
```


The Megabenthos Rpath category has `r dim(megaben)[1]` food habits database species codes:

```{r}
datatable(megaben, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Megabenthos species names and codes",
          options = list(pageLength = 10,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))
```

Which of these show up most often in predator stomachs?

```{r}
# object is called `allfh`
load(url("https://github.com/NOAA-EDAB/forageindex/raw/main/fhdat/allfh.Rdata"))

#object is called allfh21
load(url("https://github.com/NOAA-EDAB/forageindex/raw/main/fhdat/allfh21.RData"))

allfh <- allfh %>%
  dplyr::bind_rows(allfh21)

macrobenfh <- allfh %>%
  dplyr::left_join(macroben %>% 
                     dplyr::select(PYNAM, RPATH) %>%
                     setNames(., tolower(names(.)))) %>%
  dplyr::filter(!is.na(rpath))

megabenfh <- allfh %>%
  dplyr::left_join(megaben %>% 
                     dplyr::select(PYNAM, RPATH) %>%
                     setNames(., tolower(names(.)))) %>%
  dplyr::filter(!is.na(rpath))

macropreycount <- macrobenfh %>%
   #group_by(pdcomnam, pynam) %>%
   group_by(pynam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))
   #pivot_wider(names_from = pdcomnam, values_from = count) 

megapreycount <- megabenfh %>%
   #group_by(pdcomnam, pynam) %>%
   group_by(pynam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))
   #pivot_wider(names_from = pdcomnam, values_from = count) 

datatable(macropreycount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of stomachs with each macrobenthos species across all predators, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

datatable(megapreycount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of stomachs with each megabenthos species across all predators, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

```


## Which predators?

We have a couple of options for a predator list. First is to use every predator (or all above a cutoff number of observations, maybe 50 or 100 across the whole database) where something in the prey list was observed, second is to use all benthivores or some other group(s) as identified by diet similarity. 

```{r}

macropredcount <- macrobenfh %>%
  group_by(pdcomnam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))

megapredcount <- megabenfh %>%
  group_by(pdcomnam) %>%
   summarise(count = n()) %>%
   #filter(pdcomnam != "") %>%
   arrange(desc(count))

datatable(macropredcount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of macrobenthos species observations by predator, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))

datatable(megapredcount, rownames = FALSE,
          extensions = c("FixedColumns"),
          caption = "Number of megabenthos species observations by predator, NEFSC 1973-2021",
          options = list(pageLength = 25,
                         scrollX = TRUE,
                         fixedColumns = list(leftColumns = 1)))



```

The clusters based on diet similarity show several clusters that appear on the lists above. I think because we want to include so many prey groups, the cluster groupings might actually limit us, or we would have to use more than one cluster:

```{r, fig.height=12}
dietoverlap <- read_csv(here("data-raw/tgmat.2022-02-15.csv"))

# follows example here https://cran.r-project.org/web/packages/dendextend/vignettes/Cluster_Analysis.html

library(dendextend)

d_dietoverlap <- dist(dietoverlap)

guilds <- hclust(d_dietoverlap)

#plot(guilds)

dend <- as.dendrogram(guilds)

dend <- rotate(dend, 1:136)

dend <- color_branches(dend, k=6) # Brian uses 6 categories

labels(dend) <- paste(as.character(names(dietoverlap[-1]))[order.dendrogram(dend)],
                           "(",labels(dend),")", 
                           sep = "")

dend <- hang.dendrogram(dend,hang_height=0.1)

# reduce the size of the labels:
# dend <- assign_values_to_leaves_nodePar(dend, 0.5, "lab.cex")
dend <- set(dend, "labels_cex", 0.5)
# And plot:
par(mar = c(3,3,3,7))
plot(dend, 
     main = "Clustered NEFSC diet data, (complete)
     (the labels give the predator species/size)", 
     horiz =  TRUE,  nodePar = list(cex = .007))
#legend("topleft", legend = iris_species, fill = rainbow_hcl(3))

```

## Which surveys?

Start with NEFSC, find out if MA and ME/NE surveys collect diet information, and add NEAMAP later once predator and prey lists are well established.

## Spatial scale

Suggestion: use entire VAST northwest-atlantic grid for estimation and split out GOM and GB as was done for the forage fish index.

## Covariates

Suggestion: start with mean predator size at a station, number of predator species at a station, and bottom water temperature. Is there anything else we would want for benthic organisms? Reviewers suggested depth for the forage fish index.

### Fill missing physical data?

When I looked at bottom temperature data in the NEFSC survey, we had as many missing values as for surface temperature, so it is likely we will want to fill these with other data sources to avoid losing inforation. Hubert DuPontavice provided his reconstruction of bottom temperatures based on GLORYS and ROMS outputs for the NEUS. The full 1.1 GB dataset is stored locally in data-raw/bottomtemp and the identical source file is [here on google drive](https://drive.google.com/file/d/11HanY5Tzu--77HevCV5gRdNvaAvuYVYU/view?usp=share_link).

# Modeling

## VAST model setup

Univariate or multivariate? First lets try separate univariate models for each of the prey groups, one model for macrobenthos and one for megabenthos. Then we can get fancier trying to look at a multivariate model?

Do we want to separate seasons or try an annual model? I guess it depends on how the Rpaths are set up.

## VAST model selection

## VAST model bias correction and results

